---
title: "Barro Colorado Island"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{bci}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(spatstat)
library(plot.matrix)
remove(list = ls())

source("../R/get_bci.R")

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the Barro Colorado Island (BCI) dataset.
We begin by loading the data with only the most prevalent species.

```{r}
number_of_species <- 20
bci <- get_bci(least_prevalent = 50, most_prevalent = 50 + number_of_species - 1)
configuration <- bci$configuration
window <- bci$window
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

The BCI dataset also contains a series of environmental covariates.
The easiest to obtain are the elevation level and the elevation gradient, since they are included in `spatstat`.

```{r}
covariates <- list(elevation = spatstat.data::bei.extra$elev, 
                   gradient = spatstat.data::bei.extra$grad)
```

Plotting covariates maps is easy in `spatstat`.

```{r}
plot(covariates$elevation)
plot(covariates$gradient)
```

The matrix `short_range` defined below models short range interaction radii within a species (on the diagonal), and between species (outside the diagonal).
One could play around with different interaction radii, but any homogeneous interaction radius of less than $10\,\mathrm{m}$ tends to work well.

```{r}
short_range <- matrix(5, number_of_species, number_of_species)
medium_range <- matrix(10, number_of_species, number_of_species)
long_range <- matrix(30, number_of_species, number_of_species)
```

Fitting the model to the dataset is then quite easy.

```{r}
tm <- Sys.time()
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      covariates = covariates, 
                      model = "square_exponential", 
                      medium_range_model = "square_exponential", 
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      fitting_package = "glm")
print(Sys.time() - tm)
print(fit$coefficients)
par(mar = c(5.1, 5.1, 4.1, 4.1))
plot(fit$coefficients$alpha[[1]])
plot(fit$coefficients$gamma)
```