---
title: "Helianthemum dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{helianthemum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(ecespa)
library(plot.matrix)
remove(list = ls())

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the `Helianthemum` dataset from `ecespa`.
Locations of H. squamatum adult plants and seedlings in a 6 m x 7 m plot over gypsum soil in Chinchón (near Madrid, Spain). 
These are part of the data collected by Romao (2003) that have been analyzed several times (Escudero et al.2005, De la Cruz 2006, De la Cruz et al. in press.). 
The coordinates of the plans are given in cm.

```{r}
configuration <- Configuration(Helianthemum)
window <- Rectangle_window(c(2, 600), c(2, 699))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
print(configuration)
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

We provide a series of ranges for the interaction radii, and let the fitting function calibrate the model.

```{r}
short_range <- c(0, 50)
medium_range <- c(0, 50)
long_range <- c(0, 50)
model <- "Geyer"
medium_range_model <- "Geyer"
saturation <- 2
max_points <- 1000
steps <- 100000
```

We can now call the fitting function.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      fitting_package = "glm",
                      saturation = saturation)
print(summary(fit))
print(fit$coefficients)
par(mar=c(5.1, 5.1, 4.1, 4.1))
plot(fit$coefficients$alpha)
plot(fit$coefficients$gamma)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1, 
#                 mark = mean(get_marks(configuration)),
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 beta0 = parameters$beta0, 
#                 beta = matrix(0, 1, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation,
#                 max_points = max_points) 
```

It is also possible to draw from the model.

```{r}
draw <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(ppjsdm::marks(configuration)), max(ppjsdm::marks(configuration))),
                       steps = steps)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)
```
