---
title: "Helianthemum dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{helianthemum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ggplot2)
library(ppjsdm)
library(ecespa)

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the `Helianthemum` dataset from `ecespa`.
Locations of H. squamatum adult plants and seedlings in a 6 m x 7 m plot over gypsum soil in Chinchón (near Madrid, Spain). 
These are part of the data collected by Romao (2003) that have been analyzed several times (Escudero et al.2005, De la Cruz 2006, De la Cruz et al. in press.). 
The coordinates of the plans are given in cm.

```{r}
# Remove duplicate points
df <- unique(data.frame(x = Helianthemum$x,
                        y = Helianthemum$y,
                        types = Helianthemum$marks))
levels(df$types)[levels(df$types) == "adultHS"] <- "adult"
levels(df$types)[levels(df$types) == "deadpl"] <- "dead"
levels(df$types)[levels(df$types) == "survpl"] <- "alive"
configuration <- Configuration(x = df$x, y = df$y, types = df$types)
window <- Rectangle_window(c(2, 600), c(2, 699))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
print(configuration)
plot(configuration, window = window)
```

In this vignette, we shall explain how one might choose the right interaction potentials.

# Decomposing the potential into Geyer potentials

Geyer potentials are so-called hard-core potentials.
They allow us to reconstruct the true potential in terms of step functions.
In order to demonstrate the technique, we search below for interactions in terms of multiple Geyer potentials.

```{r}
steps <- 4 # Total number of Geyer potentials
nspecies <- nlevels(configuration$types)
fit <- gibbsm(configuration,
              window = window, 
              saturation = 50,
              fitting_package = "glm",
              dummy_distribution = "stratified",
              min_dummy = 5e4, 
              model = rep("Geyer", steps), # We want each of the potentials to be a Geyer potential,
                                           # with corresponding distance varying between 0 and 50 cm, see below.
              short_range = lapply(seq(from = 0, to = 50, length.out = steps + 1)[-1], 
                                   function(i) matrix(i, nspecies, nspecies)))

plot(potentials(fit, "adult", "adult"))
plot(potentials(fit, "dead", "dead"))
plot(potentials(fit, "alive", "alive"))

plot(potentials(fit, "adult", "alive"))
plot(potentials(fit, "adult", "dead"))
plot(potentials(fit, "alive", "dead"))
```

```{r}
nspecies <- nlevels(configuration$types)
fit <- gibbsm(configuration,
              window = window, 
              saturation = 10,
              fitting_package = "glm",
              dummy_distribution = "stratified",
              min_dummy = 5e4, 
              model = list("exponential", "square_exponential"),
              short_range = list(matrix(5, nspecies, nspecies),
                                 matrix(50, nspecies, nspecies)))
summary(fit)
plot(potentials(fit, 1, 1))
plot(potentials(fit, 2, 2))
plot(potentials(fit, 3, 3))

plot(potentials(fit, 1, 2))
plot(potentials(fit, 1, 3))
plot(potentials(fit, 2, 3))
```

# Minimising AIC

```{r}
saturation <- 10
nsteps <- 20
nspecies <- nlevels(configuration$types)
maxr <- 200
to_optimize <- function(df) {
  lapply(seq_len(nrow(df)), function(i) {
    set.seed(1)
    sh1 <- matrix(df$short_range[i], nspecies, nspecies)
    colnames(sh1) <- rownames(sh1) <- levels(configuration$types)
    sh1["adult", "adult"] <- df$adult_short_range[i]
    
    sh2 <- matrix(df$medium_range[i], nspecies, nspecies)
    colnames(sh2) <- rownames(sh2) <- levels(configuration$types)
    sh2["adult", "adult"] <- df$adult_short_range[i]
    fit <- ppjsdm::gibbsm(configuration, 
                          window = window, 
                          model = list(df$model[i], df$medium_range_model[i]),
                          short_range = list(sh1, sh2),
                          saturation = saturation,
                          nthreads = 4,
                          dummy_distribution = "stratified", 
                          min_dummy = 500, fitting_package = "glm")
    list(aic = fit$aic, 
         alpha1 = fit$coefficients$alpha[[1]], 
         alpha2 = fit$coefficients$alpha[[2]])
  })
}
possible_adult_short_range <- seq(from = 0, to = maxr, length.out = nsteps + 1)[-1]
possible_short_range <- seq(from = 0, to = maxr, length.out = nsteps + 1)[-1]

possible_adult_medium_range <- seq(from = 0, to = maxr, length.out = nsteps + 1)[-1]
possible_medium_range <- seq(from = 0, to = maxr, length.out = nsteps + 1)[-1]
possible_model <- c("bump")
possible_medium_range_model <- c("square_exponential")
df <- expand.grid(short_range = possible_short_range,
                  adult_short_range = possible_adult_short_range,
                  medium_range = possible_medium_range, 
                  adult_medium_range = possible_adult_medium_range,
                  model = possible_model,
                  medium_range_model = possible_medium_range_model)
df <- df[df$short_range <= df$medium_range, ]
ret <- to_optimize(df)

plot_for_two_species <- function(sp1, sp2) {
  df$aic <- sapply(ret, function(r) r$aic)
  df$alpha1 <- sapply(ret, function(r) r$alpha1[sp1, sp2])
  df$alpha2 <- sapply(ret, function(r) r$alpha2[sp1, sp2])
  
  df <- df[order(df$aic), ]
  
  x <- seq(from = 0, to = maxr, length.out = 1e3)
  g <- ggplot() + 
    geom_line(aes(x = x, y = 0), colour = "red") +
    theme_minimal() +
    xlab(NULL) +
    ylab(NULL) +
    theme(legend.title = element_blank())
  ntot <- 20
  
  for(i in seq_len(ntot)) {
    sh1 <- matrix(df[i, ]$short_range, nspecies, nspecies)
    colnames(sh1) <- rownames(sh1) <- levels(configuration$types)
    sh1["adult", "adult"] <- df[i, ]$adult_short_range
    
    sh2 <- matrix(df[i, ]$medium_range, nspecies, nspecies)
    colnames(sh2) <- rownames(sh2) <- levels(configuration$types)
    sh2["adult", "adult"] <- df[i, ]$adult_medium_range
    
    short_range <- sh1[sp1, sp2]
    medium_range <- sh2[sp1, sp2]
    potential <- function(model, r) {
      if(model == "square_bump") {
        function(x) 1 - exp(-r * r * log(2) / (x * x))
      } else if(model == "square_exponential") {
        function(x) exp(-x^2 * log(2) / (r * r))
      } else if(model == "bump") {
        function(x) 1 - exp(-r * log(2) / x)
      } else if(model == "exponential") {
        function(x) exp(-x * log(2) / r)
      }
    }
    assign(paste0("dat", i), data.frame(x = x, 
                                        short = df$alpha1[i] * potential(df[i, ]$model, r = short_range)(x),
                                        medium = df$alpha2[i] * potential(df[i, ]$medium_range_model, r = medium_range)(x),
                                        total = df$alpha1[i] * potential(df[i, ]$model, r = short_range)(x) + 
                                          df$alpha2[i] * potential(df[i, ]$medium_range_model, r = medium_range)(x)))
    g <- g +
      geom_line(data = get(paste0("dat", i)), aes(x = x, y = total, colour = "Total"), size = 1, alpha = 0.8 - 0.6 * i / ntot) +
      geom_line(data = get(paste0("dat", i)), aes(x = x, y = short, colour = "Short"), alpha = 0.2) +
      geom_line(data = get(paste0("dat", i)), aes(x = x, y = medium, colour = "Medium"), alpha = 0.2)
  }
  g
}
plot_for_two_species("adult", "adult")
plot_for_two_species("dead", "dead")
plot_for_two_species("alive", "alive")

plot_for_two_species("adult", "alive")
plot_for_two_species("adult", "dead")
plot_for_two_species("alive", "dead")
```


```{r}
configuration_ppp <- ppp(x = configuration$x,
                         y = configuration$y,
                         window = owin(window$x_range, window$y_range),
                         marks = configuration$types)

plot(envelope(configuration_ppp, fun = pcfcross, i = "adult", j = "adult", normpower = 2, r = seq(from = 0, to = 200, length.out = 1e4), nsim = 200), ylim = c(0, 10))

plot(envelope(configuration_ppp, fun = pcfcross, i = "alive", j = "alive", normpower = 2, r = seq(from = 0, to = 200, length.out = 1e4), nsim = 200), ylim = c(0, 10))

plot(envelope(configuration_ppp, fun = pcfcross, i = "dead", j = "dead", normpower = 2, r = seq(from = 0, to = 200, length.out = 1e4), nsim = 200), ylim = c(0, 10))

plot(envelope(configuration_ppp, fun = pcfcross, i = "adult", j = "alive", normpower = 2, r = seq(from = 0, to = 200, length.out = 1e4), nsim = 200), ylim = c(0, 10))

plot(envelope(configuration_ppp, fun = pcfcross, i = "adult", j = "dead", normpower = 2, r = seq(from = 0, to = 200, length.out = 1e4), nsim = 200), ylim = c(0, 10))

plot(envelope(configuration_ppp, fun = pcfcross, i = "alive", j = "dead", normpower = 2, r = seq(from = 0, to = 200, length.out = 1e4), nsim = 200), ylim = c(0, 10))
```

```{r}
saturation <- 50
dummy_factor <- 4
nspecies <- nlevels(configuration$types)

# Optimisation functions work with a function of a series of parameters,
# function below converts back to matrix form
npar <- 10
par_to_matrix <- function(par) {
  # Avoid invalid values of the parameters, while preserving continuity
  par <- abs(par)
  
  # Coeff used to avoid same values for medium and short ranges, leads to degenerate results
  # especially when both potentials have the same shape
  coef <- 1.1
  
  sh1 <- matrix(par[1], nspecies, nspecies)
  colnames(sh1) <- rownames(sh1) <- levels(configuration$types)
      
  sh2 <- matrix(coef * par[1] + par[2], nspecies, nspecies)
  colnames(sh2) <- rownames(sh2) <- levels(configuration$types)
  
  sh1["adult", "adult"] <- par[3]
  sh2["adult", "adult"] <- coef * par[3] + par[4]
  
  sh1["dead", "dead"] <- sh1["alive", "alive"] <- par[5]
  sh2["dead", "dead"] <- sh2["alive", "alive"] <- coef * par[5] + par[6]
  
  sh1["adult", "alive"] <- sh1["alive", "adult"] <- par[7]
  sh2["adult", "alive"] <- sh2["alive", "adult"] <- coef * par[7] + par[8]
  
  sh1["adult", "dead"] <- sh1["dead", "adult"] <- par[9]
  sh2["adult", "dead"] <- sh2["dead", "adult"] <- coef * par[9] + par[10]
  
  list(sh1, sh2)
}

to_minimise <- function(par, model, medium_range_model) {
  set.seed(1)
  tryCatch(
  {
    ppjsdm::gibbsm(configuration, 
                   window = window, 
                   model = list(model, medium_range_model),
                   short_range = par_to_matrix(par),
                   saturation = saturation,
                   nthreads = 4,
                   dummy_distribution = "stratified", 
                   dummy_factor = dummy_factor, 
                   fitting_package = "glm")$aic
  }, warning = function(w) {
    1e10
  })
}

possible_model <- c("bump", "exponential", "square_bump", "square_exponential")
possible_medium_range_model <- c("bump", "exponential", "square_bump", "square_exponential")
models <- expand.grid(model = possible_model,
                      medium_range_model = possible_medium_range_model)
# models <- models[!duplicated(t(apply(models, 1, sort))), ]

tm <- Sys.time()
opts <- lapply(seq_len(nrow(models)), function(i) {
  # Get initial guess by drawing 100 quasi-random draws
  draws <- abs(15 + 10 * randtoolbox::sobol(100, dim = npar, normal = TRUE))
  z <- lapply(seq_len(nrow(draws)), function(j) {
    set.seed(j)
    par <- draws[j, ]
    m <- to_minimise(par, model = models$model[i], medium_range_model = models$medium_range_model[i])
    list(par = par, min = m)
  })
  
  # Figure out which one of the draws was best
  best <- which.min(sapply(z, function(x) x$m))
  
  # And use that to initialise the optim call
  optim(par = z[[best]]$par, 
        fn = to_minimise, 
        method = "L-BFGS", # Try BFGS? CG? Nelder–Mead?
        control = list(maxit = 30000),
        model = models$model[i],
        lower = rep(0, npar),
        upper = rep(100, npar),
        medium_range_model = models$medium_range_model[i])
})
Sys.time() - tm

best_opt <- order(sapply(opts, function(o) o$value))[1]
model <- models$model[best_opt]
medium_range_model <- models$medium_range_model[best_opt]
opt <- opts[[best_opt]]

set.seed(1)
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = list(model, medium_range_model),
                      short_range = par_to_matrix(opt$par),
                      saturation = saturation,
                      nthreads = 4,
                      dummy_distribution = "stratified", 
                      min_dummy = 1e5, fitting_package = "glm")
plot(potentials(fit, "adult", "adult"))
plot(potentials(fit, "alive", "alive"))
plot(potentials(fit, "dead", "dead"))

plot(potentials(fit, "adult", "alive"))
plot(potentials(fit, "adult", "dead"))
plot(potentials(fit, "alive", "dead"))
```

```{r, fig.height = 10, fig.width = 10}
nsim <- 8
steps <- 1e5
draws <- ppjsdm::rgibbs(fit, nsim = nsim, steps = steps, starting_configuration = configuration)
plots <- lapply(draws, plot)
plots[[length(plots) + 1]] <- plot(configuration)
indices <- sample(seq_len(length(plots)))
cowplot::plot_grid(plotlist = plots[indices])
which(indices == length(indices))
```