---
title: "Kimboto dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{kimboto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(spatstat)
remove(list = ls())

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the kimboto dataset from `spatstat`.
We begin by loading the data with all species.

```{r}
print(paracou)
configuration <- Configuration(paracou)
window <- Rectangle_window(c(0, 400.8568), c(0, 524.4037))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

We provide a series of ranges for the interaction radii, and let the fitting function calibrate the model.

```{r}
short_range <- c(0, 20)
medium_range <- c(0, 20)
long_range <- c(0, 20)
model <- "square_exponential"
medium_range_model <- "square_exponential"
```

We can now call the fitting function.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      use_glmnet = TRUE)
plot(fit$complete)
print(fit$coefficients)
print(fit$best_short)
print(fit$best_medium)
print(fit$best_long)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- get_parameters_from_fit(fit)
lambda <- parameters$lambda
alpha <- parameters$alpha
gamma <- parameters$gamma
plot_papangelou(window = window,
                configuration = configuration,
                type = 1, 
                model = model, 
                medium_range_model = medium_range_model, 
                alpha = alpha, 
                lambda = lambda, 
                beta = matrix(0, 2, 0), 
                gamma = gamma, 
                covariates = list(), 
                short_range = fit$best_short, 
                medium_range = fit$best_medium, 
                long_range = fit$best_long, 
                saturation = 2) 

plot_papangelou(window = window,
                configuration = configuration,
                type = 2, 
                model = model, 
                medium_range_model = medium_range_model, 
                alpha = alpha, 
                lambda = lambda, 
                beta = matrix(0, 2, 0), 
                gamma = gamma, 
                covariates = list(), 
                short_range = fit$best_short, 
                medium_range = fit$best_medium, 
                long_range = fit$best_long, 
                saturation = 2) 
```

It is also possible to draw from the model.

```{r}
parameters <- get_parameters_from_fit(fit)
lambda <- parameters$lambda
alpha <- parameters$alpha
gamma <- parameters$gamma
draw <- ppjsdm::rgibbs(window = window, 
                       alpha = alpha, 
                       lambda = lambda, 
                       gamma = gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = fit$best_short, 
                       medium_range = fit$best_medium, 
                       long_range = fit$best_long, 
                       types = levels(types(configuration)),
                       mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))),
                       steps = 10000000)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)