---
title: "Kimboto dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{kimboto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(spatstat)
library(plot.matrix)
remove(list = ls())

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the kimboto dataset from `spatstat`.
This dataset is a point pattern of adult and juvenile Kimboto trees (Pradosia cochlearia or P. ptychandra) recorded at Paracou in French Guiana. 
See Flores (2005).

The dataset paracou is a point pattern (object of class "ppp") containing the spatial coordinates of each tree, marked by age (a factor with levels adult and juvenile. 
The survey region is a rectangle approximately 400 by 525 metres. 
Coordinates are given in metres.
We begin by loading the data with all species.

```{r}
print(paracou)
configuration <- Configuration(paracou)
window <- Rectangle_window(c(0, 400.8568), c(0, 524.4037))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```
  
We provide a series of ranges for the interaction radii, and let the fitting function calibrate the model.

```{r}
short_range <- c(0, 20)
medium_range <- c(0, 20)
long_range <- c(0, 20)
model <- "square_exponential"
medium_range_model <- "square_exponential"
steps <- 100000
max_points <- 1000
saturation <- 2
```

We can now call the fitting function.

```{r}
fit <- suppressWarnings(ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      fitting_package = "glm"))
print(summary(fit))
print(fit$coefficients)
par(mar=c(5.1, 5.1, 4.1, 4.1))
plot(fit$coefficients$alpha)
plot(fit$coefficients$gamma)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1, 
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 lambda = parameters$lambda, 
#                 beta = matrix(0, 2, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation,
#                 max_points = max_points) 
# 
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 2, 
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 lambda = parameters$lambda, 
#                 beta = matrix(0, 2, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation,
#                 max_points = max_points) 
```

It is also possible to draw from the model.

```{r}
draw <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))),
                       saturation = saturation,
                       steps = steps)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)