---
title: "Fithian et al. (2014) NSW"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{nsw}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(maptools)
library(ppjsdm)
library(raster)
library(sf)
library(spatstat)
remove(list = ls())

source("../R/get_nsw.R")

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the NSW dataset from Fithian et al. (2014).
We begin by loading the data with only the most prevalent species.

```{r}
number_of_species <- 2 # Includes the most prevalent species from the plot

nsw <- get_nsw(prevalent = number_of_species)
configuration <- nsw$configuration
window <- nsw$window
covariates <- nsw$covariates
```

Some of the covariates give `NA` values at some locations in the configuration, so we have to remove these points

```{r}
for(covariate in append(covariates, list(window))) {
  ok <- inside.owin(x_coordinates(configuration), y_coordinates(configuration), covariate)
  configuration <- Configuration(x_coordinates(configuration)[ok], 
                                 y_coordinates(configuration)[ok],
                                 types(configuration)[ok])
}
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
print(configuration)
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

The matrix `radii` defined below models interaction radii within a species (on the diagonal), and between species (outside the diagonal).

```{r}
short_range <- matrix(0.0005, number_of_species, number_of_species)
medium_range <- diag(0.0005, number_of_species)
long_range <- medium_range + diag(0.001, number_of_species)

model <- "Geyer"
medium_range_model <- "Geyer"
```

Fitting the model to the dataset is then quite easy.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      covariates = covariates, 
                      model = model, 
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      use_glmnet = TRUE)
plot(fit$complete)
plot(fit$cv)
```
