---
title: "Queensland rainforest"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{qld}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(spatstat)
library(plot.matrix)
remove(list = ls())

source("../R/get_qld.R")

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the Queensland rainforest dataset from CSIRO.
We begin by loading the data with the most prevalent species.

```{r}
index_of_plot <- 3 # Between 1 and 20
year <- 2011 # Year of census
number_of_species <- 2 # Includes the most prevalent species from the plot

qld <- get_qld(index = index_of_plot,
               year = year, 
               prevalent = number_of_species)
configuration <- qld$configuration
window <- qld$window
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

The function `gibbsm` fits a multivariate Gibbs point process to our dataset.
For example,

```{r}
short_range <- c(0, 10)
medium_range <- c(0, 10)
long_range <- c(0, 10)

fit <- ppjsdm::gibbsm(configuration, 
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      window = window, 
                      use_glmnet = FALSE)
print(fit$coefficients)
par(mar = c(5.1, 5.1, 4.1, 4.1))
plot(fit$coefficients$alpha)
plot(fit$coefficients$gamma)
print(fit$aic)
print(fit$bic)
```

It is then possible to draw from the model, as can be seen below.

```{r}
# parameters <- fit$coefficients
# draw <- ppjsdm::rgibbs(window = window, 
#                        alpha = parameters$alpha, 
#                        lambda = parameters$lambda, 
#                        gamma = parameters$gamma,
#                        short_range = parameters$short_range,
#                        medium_range = parameters$medium_range,
#                        long_range = parameters$long_range,
#                        types = levels(types(configuration)))
# print(draw)
# 
# par(mar = c(5, 4, 4, 13) + 0.1)
# plot(draw, window = window)
```

Let us increase the number of species accounted for.

```{r}
saturation <- 2
max_points <- 1000

index_of_plot <- 3 # Between 1 and 20
year <- 2011 # Year of census
number_of_species <- 5 # Includes the most prevalent species from the plot

qld <- get_qld(index = index_of_plot,
               year = year, 
               prevalent = number_of_species)
configuration <- qld$configuration
window <- qld$window

par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)

short_range <- c(0, 10)
medium_range <- c(0, 10)
long_range <- c(0, 10)

model <- "square_bump"
medium_range_model <- "square_exponential"

fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      model = model,
                      medium_range_model = medium_range_model,
                      use_glmnet = FALSE)
print(fit$coefficients)
par(mar = c(5.1, 5.1, 4.1, 4.1))
plot(fit$coefficients$alpha)
plot(fit$coefficients$gamma)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 2, 
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 lambda = parameters$lambda, 
#                 beta = matrix(0, number_of_species, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation,
#                 max_points = max_points) 
```