---
title: "Seedlings dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{seedlings1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(ecespa)
remove(list = ls())

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the `seedlings1` dataset from `ecespa`.
This dataset is a marked point pattern of seedlings of H. squamatumgrowing in a gyp-sophylous plant community in Central Spain.
The datasets contains the locations of the seedlings marked with their heigth. 
Both the coordinates and the heigth of the seedlings are given in cm.

# Taking marks into account

If marks (i.e. their height) are provided, the interaction radii are proportional to the marks.
We begin with that setting.

```{r}
configuration <- Configuration(seedlings1$x, seedlings1$y, marks = seedlings1$marks)
window <- Rectangle_window(c(0, 325), c(0, 300))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
print(configuration)
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

We provide a series of ranges for the interaction radii, and let the fitting function calibrate the model.

```{r}
short_range <- c(0, 20)
medium_range <- c(0, 20)
long_range <- c(0, 20)
model <- "square_exponential"
medium_range_model <- "square_exponential"
saturation <- 2
steps <- 100000
```

We can now call the fitting function.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      use_glmnet = FALSE,
                      use_aic = TRUE,
                      saturation = saturation)
print(summary(fit))
print(fit$coefficients)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1, 
#                 mark = mean(get_marks(configuration)),
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 beta0 = parameters$beta0, 
#                 beta = matrix(0, 1, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation) 
```

It is also possible to draw from the model.

```{r}
draw <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))),
                       saturation = saturation,
                       steps = steps)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)
```

# Interaction radii not proportional to marks

In this section, we disregard the height of the seedlings, and consider interaction radii in centimetres.

```{r}
configuration <- Configuration(seedlings1$x, seedlings1$y)
```

We call the fitting function on this unmarked point process.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      use_glmnet = FALSE,
                      use_aic = TRUE,
                      saturation = saturation)
print(summary(fit))
print(fit$coefficients)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1,
#                 mark = mean(get_marks(configuration)),
#                 model = model,
#                 medium_range_model = medium_range_model,
#                 alpha = parameters$alpha,
#                 beta0 = parameters$beta0,
#                 beta = matrix(0, 1, 0),
#                 gamma = parameters$gamma,
#                 covariates = list(),
#                 short_range = parameters$short_range,
#                 medium_range = parameters$medium_range,
#                 long_range = parameters$long_range,
#                 saturation = saturation)
```

And as previously, we draw from the model.

```{r}
draw <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))),
                       saturation = saturation,
                       steps = steps)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)
```
