---
title: "Norwegian spruces dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{spruces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(spatstat)
remove(list = ls())

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the `spruces` dataset from `spatstat`.
The data give the locations of Norwegian spruce trees in a natural forest stand in Saxonia, Germany.
Each tree is marked with its diameter at breast height.

# Taking marks into account

If marks are provided, the interaction radii are proportional to the marks, i.e. their diameter at breast height.
We begin with that setting.

```{r}
configuration <- Configuration(spruces$x, spruces$y, marks = spruces$marks)
window <- Rectangle_window(c(0, 56), c(0, 38))
spatstat_window <- owin(c(0, 56), c(0, 38))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
print(configuration)
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

We provide a series of ranges for the interaction radii, and let the fitting function calibrate the model.

```{r}
short_range <- c(0, 20)
medium_range <- c(0, 20)
long_range <- c(0, 20)
model <- "square_exponential"
medium_range_model <- "square_exponential"
saturation <- 2
steps <- 100000
```

We can now call the fitting function.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      use_glmnet = FALSE,
                      use_aic = TRUE,
                      saturation = saturation)
print(summary(fit))
print(fit$coefficients)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1, 
#                 mark = mean(get_marks(configuration)),
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 beta0 = parameters$beta0, 
#                 beta = matrix(0, 1, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation,
#                 max_points = max_points) 
```

It is also possible to draw from the model.

```{r}
draw <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))),
                       saturation = saturation,
                       steps = steps)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)
```

A more standard approach to check goodness of fit is a comparison of K functions.

```{r}
env <- envelope.ppp(ppp(configuration$x, configuration$y, window = spatstat_window), simulate = function(X) {
  z <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))),
                       saturation = saturation,
                       steps = steps)
  ppp(z$x, z$y, window = spatstat_window)
  })
plot(env)
```

# Interaction radii not proportional to marks

In this section, we disregard the diameter at breast height, and consider interaction radii in metres.

```{r}
configuration <- Configuration(spruces$x, spruces$y)
```

We call the fitting function on this unmarked point process.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                        use_glmnet = FALSE,
                      use_aic = TRUE,
                      saturation = saturation)
print(summary(fit))
print(fit$coefficients)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1, 
#                 mark = mean(get_marks(configuration)),
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 beta0 = parameters$beta0, 
#                 beta = matrix(0, 1, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation) 
```

And as previously, we draw from the model.

```{r}
draw <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))),
                       saturation = saturation,
                       steps = steps)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)
```

And let us also plot the K function.

```{r}
env <- envelope.ppp(ppp(configuration$x, configuration$y, window = spatstat_window), simulate = function(X) {
  z <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))),
                       saturation = saturation,
                       steps = steps)
  ppp(z$x, z$y, window = spatstat_window)
  })
plot(env)
```

<!-- # Comparison to classical Hard-core model. -->

<!-- ```{r} -->
<!-- configuration <- Configuration(spruces$x, spruces$y) -->
<!-- model <- "Geyer" -->
<!-- medium_range_model <- "Geyer" -->
<!-- ``` -->

<!-- We call the fitting function on this unmarked point process. -->

<!-- ```{r} -->
<!-- fit <- ppjsdm::gibbsm(configuration,  -->
<!--                       window = window,  -->
<!--                       model = model, -->
<!--                       medium_range_model = medium_range_model, -->
<!--                       short_range = short_range, -->
<!--                       medium_range = medium_range, -->
<!--                       long_range = long_range, -->
<!--                       use_glmnet = FALSE, -->
<!--                       use_aic = TRUE, -->
<!--                       saturation = saturation) -->
<!-- print(summary(fit)) -->
<!-- print(fit$coefficients) -->
<!-- print(fit$aic) -->
<!-- print(fit$bic) -->
<!-- ``` -->

<!-- We may then plot the corresponding Papangelou conditional intensity. -->

<!-- ```{r} -->
<!-- parameters <- fit$coefficients -->
<!-- # plot_papangelou(window = window, -->
<!-- #                 configuration = configuration, -->
<!-- #                 type = 1,  -->
<!-- #                 mark = mean(get_marks(configuration)), -->
<!-- #                 model = model,  -->
<!-- #                 medium_range_model = medium_range_model,  -->
<!-- #                 alpha = parameters$alpha,  -->
<!-- #                 beta0 = parameters$beta0,  -->
<!-- #                 beta = matrix(0, 1, 0),  -->
<!-- #                 gamma = parameters$gamma,  -->
<!-- #                 covariates = list(),  -->
<!-- #                 short_range = parameters$short_range,  -->
<!-- #                 medium_range = parameters$medium_range,  -->
<!-- #                 long_range = parameters$long_range,  -->
<!-- #                 saturation = saturation)  -->
<!-- ``` -->

<!-- And as previously, we draw from the model. -->

<!-- ```{r} -->
<!-- draw <- ppjsdm::rgibbs(window = window, -->
<!--                        alpha = parameters$alpha, -->
<!--                        beta0 = parameters$beta0, -->
<!--                        gamma = parameters$gamma, -->
<!--                        model = model, -->
<!--                        medium_range_model = medium_range_model, -->
<!--                        short_range = parameters$short_range, -->
<!--                        medium_range = parameters$medium_range, -->
<!--                        long_range = parameters$long_range, -->
<!--                        types = levels(types(configuration)), -->
<!--                        mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))), -->
<!--                        saturation = saturation, -->
<!--                        steps = steps) -->
<!-- print(draw) -->

<!-- par(mar = c(5, 4, 4, 13) + 0.1) -->
<!-- plot(draw, window = window) -->
<!-- ``` -->

<!-- And let us finally plot the K function. -->

<!-- ```{r} -->
<!-- env <- envelope.ppp(ppp(configuration$x, configuration$y, window = spatstat_window), simulate = function(X) { -->
<!--   z <- ppjsdm::rgibbs(window = window, -->
<!--                        alpha = parameters$alpha, -->
<!--                        beta0 = parameters$beta0, -->
<!--                        gamma = parameters$gamma, -->
<!--                        model = model, -->
<!--                        medium_range_model = medium_range_model, -->
<!--                        short_range = parameters$short_range, -->
<!--                        medium_range = parameters$medium_range, -->
<!--                        long_range = parameters$long_range, -->
<!--                        types = levels(types(configuration)), -->
<!--                        mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))), -->
<!--                        saturation = saturation, -->
<!--                        steps = steps) -->
<!--   ppp(z$x, z$y, window = spatstat_window) -->
<!--   }) -->
<!-- plot(env) -->
<!-- ``` -->