---
title: "Norwegian spruces dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{spruces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(spatstat)
remove(list = ls())

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the `spruces` dataset from `spatstat`.
The data give the locations of Norwegian spruce trees in a natural forest stand in Saxonia, Germany.
Each tree is marked with its diameter at breast height.

# Taking marks into account

If marks are provided, the interaction radii are proportional to the marks, i.e. their diameter at breast height.
We begin with that setting.

```{r}
configuration <- Configuration(spruces$x, spruces$y, marks = spruces$marks)
window <- Rectangle_window(c(0, 56), c(0, 38))
spatstat_window <- owin(c(0, 56), c(0, 38))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
print(configuration)
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

We provide a series of ranges for the interaction radii, and let the fitting function calibrate the model.

```{r}
short_range <- c(0, 20)
medium_range <- c(0, 20)
long_range <- c(0, 20)
model <- "square_exponential"
medium_range_model <- "square_exponential"
saturation <- 2
steps <- 100000
```

We can now call the fitting function.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      fitting_package = "glm",
                      saturation = saturation)
print(summary(fit))
print(fit$coefficients)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1, 
#                 mark = mean(get_marks(configuration)),
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 beta0 = parameters$beta0, 
#                 beta = matrix(0, 1, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation,
#                 max_points = max_points) 
```

It is also possible to draw from the model.

```{r}
draw <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(ppjsdm::marks(configuration)), max(ppjsdm::marks(configuration))),
                       saturation = saturation,
                       steps = steps)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)
```

A more standard approach to check goodness of fit is a comparison of K functions.

```{r}
env <- envelope.ppp(ppp(configuration$x, configuration$y, window = spatstat_window), simulate = function(X) {
  z <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(ppjsdm::marks(configuration)), max(ppjsdm::marks(configuration))),
                       saturation = saturation,
                       steps = steps)
  ppp(z$x, z$y, window = spatstat_window)
  })
plot(env)
```

# Interaction radii not proportional to marks

In this section, we disregard the diameter at breast height, and consider interaction radii in metres.

```{r}
configuration <- Configuration(spruces$x, spruces$y)
```

We call the fitting function on this unmarked point process.

```{r}
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      fitting_package = "glm",
                      saturation = saturation)
print(summary(fit))
print(fit$coefficients)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1, 
#                 mark = mean(get_marks(configuration)),
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 beta0 = parameters$beta0, 
#                 beta = matrix(0, 1, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation) 
```

And as previously, we draw from the model.

```{r}
draw <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(ppjsdm::marks(configuration)), max(ppjsdm::marks(configuration))),
                       saturation = saturation,
                       steps = steps)
print(draw)

par(mar = c(5, 4, 4, 13) + 0.1)
plot(draw, window = window)
```

And let us also plot the K function.

```{r}
env <- envelope.ppp(ppp(configuration$x, configuration$y, window = spatstat_window), simulate = function(X) {
  z <- ppjsdm::rgibbs(window = window,
                       alpha = parameters$alpha,
                       beta0 = parameters$beta0,
                       gamma = parameters$gamma,
                       model = model,
                       medium_range_model = medium_range_model,
                       short_range = parameters$short_range,
                       medium_range = parameters$medium_range,
                       long_range = parameters$long_range,
                       types = levels(types(configuration)),
                       mark_range = c(min(ppjsdm::marks(configuration)), max(ppjsdm::marks(configuration))),
                       saturation = saturation,
                       steps = steps)
  ppp(z$x, z$y, window = spatstat_window)
  })
plot(env)
```

# Using two short-range potentials

In the latest version of `ppjsdm`, we can use multiple short-range potentials.
Let us explore how that would work.

```{r}
saturation <- 10
to_optimize <- function(df) {
  lapply(seq_len(nrow(df)), function(i) {
    set.seed(1)
    fit <- ppjsdm::gibbsm(configuration, 
                          window = window, 
                          model = list(df$model[i], df$medium_range_model[i]),
                          short_range = list(df$short_range[i], df$medium_range[i]),
                          saturation = saturation,
                          dummy_distribution = "stratified", 
                          min_dummy = 1, dummy_factor = 1e10,
                          max_dummy = 1e3, fitting_package = "glm")
    list(aic = fit$aic, 
         alpha1 = fit$coefficients$alpha[[1]], 
         alpha2 = fit$coefficients$alpha[[2]])
  })
}
possible_short_range <- seq(from = 0.1, to = 4, by = 0.1)
possible_medium_range <- seq(from = 0.1, to = 4, by = 0.1)
possible_model <- c("square_exponential", "exponential", "square_bump", "bump")
possible_medium_range_model <- c("square_exponential", "exponential", "square_bump", "bump")
df <- expand.grid(short_range = possible_short_range,
                  medium_range = possible_medium_range, 
                  model = possible_model,
                  medium_range_model = possible_medium_range_model)
df <- df[df$short_range <= df$medium_range, ]
ret <- to_optimize(df)
df$aic <- sapply(ret, function(r) r$aic)
df$alpha1 <- sapply(ret, function(r) r$alpha1)
df$alpha2 <- sapply(ret, function(r) r$alpha2)

df <- df[order(df$aic), ]

x <- seq(from = 0, to = 5, length.out = 1e3)
g <- ggplot() + 
  geom_line(aes(x = x, y = 0), colour = "red") +
  theme_minimal()
ntot <- 100

for(i in seq_len(ntot)) {
  short_range <- df[i, ]$short_range
  medium_range <- df[i, ]$medium_range
  potential <- function(model, r) {
    if(model == "square_bump") {
      function(x) 1 - exp(-r * r * log(2) / (x * x))
    } else if(model == "square_exponential") {
      function(x) exp(-x^2 * log(2) / (r * r))
    } else if(model == "bump") {
      function(x) 1 - exp(-r * log(2) / x)
    } else if(model == "exponential") {
      function(x) exp(-x * log(2) / r)
    }
  }
  assign(paste0("dat", i), data.frame(x = x, 
                                      short = df$alpha1[i] * potential(df[i, ]$model, r = short_range)(x),
                                      medium = df$alpha2[i] * potential(df[i, ]$medium_range_model, r = medium_range)(x),
                                      total = df$alpha1[i] * potential(df[i, ]$model, r = short_range)(x) + 
                                        df$alpha2[i] * potential(df[i, ]$medium_range_model, r = medium_range)(x)))
  g <- g +
    geom_line(data = get(paste0("dat", i)), aes(x = x, y = total, colour = "Total"), alpha = 0.3) +
    geom_line(data = get(paste0("dat", i)), aes(x = x, y = short, colour = "Short"), alpha = 0.2) +
    geom_line(data = get(paste0("dat", i)), aes(x = x, y = medium, colour = "Medium"), alpha = 0.2)
}
g
```

<!-- # Comparison to classical Hard-core model. -->

<!-- ```{r} -->
<!-- configuration <- Configuration(spruces$x, spruces$y) -->
<!-- model <- "Geyer" -->
<!-- medium_range_model <- "Geyer" -->
<!-- ``` -->

<!-- We call the fitting function on this unmarked point process. -->

<!-- ```{r} -->
<!-- fit <- ppjsdm::gibbsm(configuration,  -->
<!--                       window = window,  -->
<!--                       model = model, -->
<!--                       medium_range_model = medium_range_model, -->
<!--                       short_range = short_range, -->
<!--                       medium_range = medium_range, -->
<!--                       long_range = long_range, -->
<!--                       use_glmnet = FALSE, -->
<!--                       use_aic = TRUE, -->
<!--                       saturation = saturation) -->
<!-- print(summary(fit)) -->
<!-- print(fit$coefficients) -->
<!-- print(fit$aic) -->
<!-- print(fit$bic) -->
<!-- ``` -->

<!-- We may then plot the corresponding Papangelou conditional intensity. -->

<!-- ```{r} -->
<!-- parameters <- fit$coefficients -->
<!-- # plot_papangelou(window = window, -->
<!-- #                 configuration = configuration, -->
<!-- #                 type = 1,  -->
<!-- #                 mark = mean(get_marks(configuration)), -->
<!-- #                 model = model,  -->
<!-- #                 medium_range_model = medium_range_model,  -->
<!-- #                 alpha = parameters$alpha,  -->
<!-- #                 beta0 = parameters$beta0,  -->
<!-- #                 beta = matrix(0, 1, 0),  -->
<!-- #                 gamma = parameters$gamma,  -->
<!-- #                 covariates = list(),  -->
<!-- #                 short_range = parameters$short_range,  -->
<!-- #                 medium_range = parameters$medium_range,  -->
<!-- #                 long_range = parameters$long_range,  -->
<!-- #                 saturation = saturation)  -->
<!-- ``` -->

<!-- And as previously, we draw from the model. -->

<!-- ```{r} -->
<!-- draw <- ppjsdm::rgibbs(window = window, -->
<!--                        alpha = parameters$alpha, -->
<!--                        beta0 = parameters$beta0, -->
<!--                        gamma = parameters$gamma, -->
<!--                        model = model, -->
<!--                        medium_range_model = medium_range_model, -->
<!--                        short_range = parameters$short_range, -->
<!--                        medium_range = parameters$medium_range, -->
<!--                        long_range = parameters$long_range, -->
<!--                        types = levels(types(configuration)), -->
<!--                        mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))), -->
<!--                        saturation = saturation, -->
<!--                        steps = steps) -->
<!-- print(draw) -->

<!-- par(mar = c(5, 4, 4, 13) + 0.1) -->
<!-- plot(draw, window = window) -->
<!-- ``` -->

<!-- And let us finally plot the K function. -->

<!-- ```{r} -->
<!-- env <- envelope.ppp(ppp(configuration$x, configuration$y, window = spatstat_window), simulate = function(X) { -->
<!--   z <- ppjsdm::rgibbs(window = window, -->
<!--                        alpha = parameters$alpha, -->
<!--                        beta0 = parameters$beta0, -->
<!--                        gamma = parameters$gamma, -->
<!--                        model = model, -->
<!--                        medium_range_model = medium_range_model, -->
<!--                        short_range = parameters$short_range, -->
<!--                        medium_range = parameters$medium_range, -->
<!--                        long_range = parameters$long_range, -->
<!--                        types = levels(types(configuration)), -->
<!--                        mark_range = c(min(get_marks(configuration)), max(get_marks(configuration))), -->
<!--                        saturation = saturation, -->
<!--                        steps = steps) -->
<!--   ppp(z$x, z$y, window = spatstat_window) -->
<!--   }) -->
<!-- plot(env) -->
<!-- ``` -->