---
title: "Swamp forest dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{swamp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ppjsdm)
library(ecespa)
library(plot.matrix)
remove(list = ls())

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the `swamp` dataset from `ecespa`.
Locations and botanical classification of trees in a plot in the Savannah River. 
Locations are given in metres, rounded to the nearest 0.1 metre. 
The data come from a 1-ha (200 m x 50 m) plot in the Savannah River Site, South Carolina, USA. 
The 734 mapped stems included 156 Carolina ash (Fraxinus caroliniana), 215 Water tupelo (Nyssa aquatica), 205 Swamp tupelo (Nyssa sylvatica), 98 Bald cypress (Taxodium distichum) and 60 stems of 8 additional species.
We begin by loading the data with all species.

```{r}
configuration <- Configuration(swamp$y, swamp$x, swamp$sp)
window <- Rectangle_window(c(0, 200), c(0, 50))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

We provide a series of ranges for the interaction radii, and let the fitting function calibrate the model.
We can then call the fitting function.

```{r}
short_range <- c(0, 20)
medium_range <- c(0, 20)
long_range <- c(0, 20)
model <- "exponential"
medium_range_model <- "square_exponential"
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      medium_range_model = medium_range_model,
                      short_range = short_range,
                      medium_range = medium_range,
                      long_range = long_range,
                      fitting_package = "glm")
fit_summary <- summary(fit)
print(fit_summary)
print(fit$coefficients)
par(mar = c(5.1, 5.1, 4.1, 4.1))
plot(fit$coefficients$alpha)
plot(fit$coefficients$gamma)
print(fit$aic)
print(fit$bic)
```

We may then plot the corresponding Papangelou conditional intensity.

```{r}
# parameters <- fit$coefficients
# plot_papangelou(window = window,
#                 configuration = configuration,
#                 type = 1, 
#                 model = model, 
#                 medium_range_model = medium_range_model, 
#                 alpha = parameters$alpha, 
#                 lambda = parameters$lambda, 
#                 beta = matrix(0, 6, 0), 
#                 gamma = parameters$gamma, 
#                 covariates = list(), 
#                 short_range = parameters$short_range, 
#                 medium_range = parameters$medium_range, 
#                 long_range = parameters$long_range, 
#                 saturation = saturation,
#                 max_points = max_points) 
```