---
title: "Swamp forest dataset"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{swamp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
                                                  "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)

```

```{r setup}
library(ecespa)
library(ggplot2)
library(ppjsdm)
library(spatstat)
remove(list = ls())

set.seed(1)
```

This vignette explains how to use the `ppjsdm` package with the `swamp` dataset from `ecespa`.
It involves multiple species and thus assumes that you have already gone through one of the simpler vignettes.

# Formatting the data

Locations and botanical classification of trees in a plot in the Savannah River. 
Locations are given in metres, rounded to the nearest 0.1 metre. 
The data come from a 1-ha (200 m x 50 m) plot in the Savannah River Site, South Carolina, USA. 
The 734 mapped stems included 156 Carolina ash (Fraxinus caroliniana), 215 Water tupelo (Nyssa aquatica), 205 Swamp tupelo (Nyssa sylvatica), 98 Bald cypress (Taxodium distichum) and 60 stems of 8 additional species.
We begin by loading the data with all species.

```{r}
configuration <- ppjsdm::Configuration(swamp$y, swamp$x, swamp$sp)
window <- ppjsdm::Rectangle_window(c(0, 200), c(0, 50))
nspecies <- length(unique(swamp$sp))
```

The point configuration is plotted below.

```{r, fig.height = 4, fig.align = 'center', fig}
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration, window = window)
```

Although we do not have any environmental covariates for this dataset, a horizontal unmeasured covariate is thought to drive the distribution.

```{r}
covariates <- list(horizontal = function(x, y) x / 100)
```

# A basic fit

We start with a first elementary fit to properly choose the number of dummy points.

```{r}
short_range <- matrix(5, nspecies, nspecies)
model <- "exponential"
ndummy <- 5e3
nthreads <- 4
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      covariates = covariates,
                      short_range = short_range,
                      fitting_package = "glm",
                      dummy_distribution = "stratified",
                      min_dummy = ndummy,
                      nthreads = nthreads,
                      dummy_factor = 1)

summary(fit)
```

# A better fit

Let us first find good choices for the short-range interaction radii.
We begin by assuming that all species have the same interaction radius.

```{r}
ndummy <- 1e3
nthreads <- 4
to_optimize <- function(df) {
  sapply(seq_len(nrow(df)), function(i) {
    set.seed(1)
    fit <- ppjsdm::gibbsm(configuration, 
                          window = window, 
                          model = df$model[i],
                          covariates = covariates,
                          short_range = matrix(df$short[i], nspecies, nspecies),
                          fitting_package = "glm",
                          dummy_distribution = "stratified",
                          min_dummy = ndummy,
                          nthreads = nthreads,
                          dummy_factor = 1)
    fit$aic
  })
}
possible_short <- seq(from = 0.5, to = 20, length.out = 50)
possible_model <- c("square_exponential", "exponential", "square_bump", "bump")
df <- expand.grid(short = possible_short, model = possible_model)
df$aic <- to_optimize(df)
df$potentials <- df$model

ggplot(df) + geom_point(aes(x = short, y = aic, colour = potentials))
```

The minimum AIC is attained around 5 meters.

```{r}
df$short[which.min(df$aic[df$potentials == "exponential"])]
```

Fixing values of 5 meters for all other interaction radii, is it possible to do better for some species?
We start from a matrix of interaction radii of 5 meters, and change one of the diagonal coefficients at a time, checking each time if we attain a lower AIC.

```{r}
ndummy <- 1e3
nthreads <- 4
short_range <- matrix(5, nspecies, nspecies)
model <- "exponential"
to_optimize <- function(df) {
  sapply(seq_len(nrow(df)), function(i) {
    short_range <- short_range
    short_range[df$sp[i], df$sp[i]] <- df$short[i]
    set.seed(1)
    fit <- ppjsdm::gibbsm(configuration, 
                          window = window, 
                          model = model,
                          covariates = covariates,
                          short_range = short_range,
                          fitting_package = "glm",
                          dummy_distribution = "stratified",
                          min_dummy = ndummy,
                          nthreads = nthreads,
                          dummy_factor = 1)
    fit$aic
  })
}
possible_short <- seq(from = 0.1, to = 7, length.out = 100)
possible_species <- seq_len(nspecies)
df <- expand.grid(short = possible_short, sp = possible_species)
df$aic <- to_optimize(df)
df$sp_name <- levels(swamp$sp)[df$sp]

ggplot(df) + geom_point(aes(x = short, y = aic, colour = sp_name))
```


We see here completely different behaviours among the five species.
The species Bald cypress (`TD`), Carolina ash (`FX`) and the other species class (`OT`) all attain a lower AIC when choosing intra-species interaction radii smaller than 1 meter.
The other two species interact at larger distances.

This procedure could also be repeated for all pairs of species.
In addition, here we have fixed some values of the interaction radii, and minimised the AIC by changing one variable at a time.
This could more rigorously be done jointly, see our manuscript for more details.
The main insights derived above still hold when doing a joint optimisation.
We find the following optimal short-range interaction radii.

```{r}
df$short[which.min(df$aic[df$sp == 1])]
df$short[which.min(df$aic[df$sp == 2])]
df$short[which.min(df$aic[df$sp == 3])]
df$short[which.min(df$aic[df$sp == 4])]
df$short[which.min(df$aic[df$sp == 5])]
```

And so we may use the following as the short-range interaction matrix.

```{r}
short_range <- matrix(5, nspecies, nspecies)
short_range[1, 1] <- 0.4
short_range[2, 2] <- 3.8
short_range[3, 3] <- 6.5
short_range[4, 4] <- 0.6
short_range[5, 5] <- 0.4
```

We can then run another fit of the model, this time increasing the number of dummy points to get the best possible fit.

```{r}
model <- "exponential"
ndummy <- 1e5
nthreads <- 4
fit <- ppjsdm::gibbsm(configuration, 
                      window = window, 
                      model = model,
                      covariates = covariates,
                      short_range = short_range,
                      fitting_package = "glm",
                      dummy_distribution = "stratified",
                      min_dummy = ndummy,
                      nthreads = nthreads,
                      dummy_factor = 1)

summary(fit)
```

The short-range interaction coefficients matrix may easily be obtained as 

```{r}
coef(fit)$alpha
```

We may then plot the corresponding Papangelou conditional intensity, for example for the second species (swamp tupelo `NS`).

```{r, fig.height = 5, fig.width = 10}
plot_papangelou(fit,
                type = 2,
                use_log = TRUE,
                grid_steps = c(2000, 500))
```

This is quite different from the fifth species (bald cypress `TD`)
This species has negative interactions with all other individuals, including those of its own class.
The species is indeed more spread out than other species in the dataset.

```{r, fig.height = 5, fig.width = 10}
plot_papangelou(fit,
                type = 5,
                use_log = TRUE,
                grid_steps = c(2000, 500))
```

# Model assessment

Let us see how well the model has performed on a specific species, say the water tupelo `NX`.
We start with a basic inhomogeneous Poisson point process fit.

```{r}
sp <- "NX"
spatstat_ppp <- ppp(x = swamp$x[swamp$sp == sp],
                    y = swamp$y[swamp$sp == sp],
                    window = owin(c(0, 50), c(0, 200)))
spatstat_fit <- ppm(spatstat_ppp ~ 1 + y)
env <- envelope(spatstat_fit, fun = Linhom, nsim = 200, 
                normpower = 2, leaveoneout = FALSE, verbose = FALSE)
plot(env)
```

The Poisson point process is not able to capture the short-range clustering in the species.
Is our model better able to model this species?
The `ppjsdm` package is able to simulate one species conditional on the others, conditional on the fitted model.
To begin answering the question, let us draw one such sample of `NS`, and see how it compares to the actual distribution.

```{r}
sp <- "NX"
sp_index <- which(names(coef(fit)$beta0) == sp)
only_species <- ppjsdm::Configuration(x = swamp$y[swamp$sp == sp], 
                                      y = swamp$x[swamp$sp == sp],
                                      types = swamp$sp[swamp$sp == sp])
everything_except_species <- ppjsdm::Configuration(x = swamp$y[swamp$sp != sp], 
                                                   y = swamp$x[swamp$sp != sp],
                                                   types = swamp$sp[swamp$sp != sp])

sample <- ppjsdm::rgibbs(fit,
                         steps = 1e4, # Number of steps in the Metropolis-Hastings algorithm
                         starting_configuration = only_species, # Starting configuration in the simulation
                         only_simulate_these_types = sp_index, # Which species to simulate
                         conditional_configuration = everything_except_species) # Conditional on these individuals
true_vs_sample <- ppjsdm::Configuration(x = c(only_species$x, sample$x),
                                        y = c(only_species$y, sample$y),
                                        types = c(rep("true", length(only_species$x)), 
                                                  rep("simulated", length(only_species$x))))
plot(true_vs_sample)
```

This looks quite convincing.
To test how well our model is performing, we need to draw a few of these samples, and compute some summary statistic for each one.
We can then test whether or not the actual distribution falls within the same range of values as those obtained by simulation.
In the code below, we use the inhomogeneous $L$-function.

```{r}
nsim <- 50
samples <- lapply(seq_len(nsim), function(i) {
  ppjsdm::rgibbs(fit,
                 steps = 1e4,
                 starting_configuration = only_species,
                 only_simulate_these_types = sp_index,
                 conditional_configuration = everything_except_species)
})

samples <- lapply(samples, function(x) ppp(x = x$x, y = x$y, window = owin(c(0, 200), c(0, 50))))

only_species_ppp <- ppp(x = swamp$y[swamp$sp == sp], 
                        y = swamp$x[swamp$sp == sp],
                        window = owin(c(0, 200), c(0, 50)))
env <- envelope(only_species_ppp, fun = Linhom, nsim = length(samples), 
                normpower = 2, leaveoneout = FALSE, simulate = samples, verbose = FALSE)
plot(env)
abline(a = 0, b = 1, col = "blue")
```

The results look much better!
The model has properly modeled the conditional distribution of the water tupelo.
